Comparing corrosion control treatments using a robust Bayesian generalized additive model

### 2022-05-05 Rmarkdown/paper-si.Rmd
preprint submission

This document contains 5 figures, 4 tables, and 10 pages.

@@

```{r figure-s5, fig.cap="**Figure S5.** In the model fitted to simulated data, **(a)** Seasonal component of the trend in dissolved lead. **(b)** Global multiyear component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) multiyear trends in dissolved and particulate lead, representing the deviations of each time series from the global multiyear trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s5.png"))
```

>>>>

This document contains 8 figures, 4 tables, and 12 pages.

@@

```{r figure-s5, fig.cap="**Figure S5.** In the model fitted to simulated data, **(a)** seasonal component of the trend in dissolved lead. **(b)** Global multiyear component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) multiyear trends in dissolved and particulate lead, representing the deviations of each time series from the global multiyear trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s5.png"))
```

### 2022-05-05 Rmarkdown/paper.Rmd
preprint submission

All of the above present barriers to reliable statistical inference, but left-censoring and autocorrelation can be particularly challenging. Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data. And while rank-based procedures such as the rank-sum test can accomodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].

@@

The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2). This is because it accounts for non-linear patterns that the simpler models ignore. GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

@@

To compare corrosion control treatments using data from the model distribution system, we fitted separate GAMs with continuous-time first-order autocorrelated errors to the particulate (>0.45 µm) and dissolved (<0.45 µm) lead time series. Particulate lead was determined as the difference between total and dissolved lead. Particulate lead concentrations less than the MDL of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored. The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration. Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were omitted.

@@

The models included one-dimensional global smooths describing multiyear variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a). Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a heirarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019]. The multiyear smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.

@@

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. The autocorrelation coeffient $\phi^s$ is defined in equation $(4)$.

@@

### Estimating derivatives

@@

The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers. Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset. Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation. The distribution of the autocorrelation coefficient $\phi$ is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.

@@

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe copuples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term.

@@

```{r figure-6, fig.cap="**Figure 6.** Time series of dissolved and particulate lead representing three orthophosphate dosing protocols (columns) and two pipe configurations (rows; the full lead pipe configuration---Pb #1--2---is duplicated). The expected value of the posterior predictive distribution (*i.e.*, $\\mu_t$ in Equation 5) is summarized by the median (blue and red lines) and the 2.5--97.5^th^ percentiles (red/blue shaded regions). The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and black ticks at the top of each plot represent values outside the plot limits. Other significant disturbances are labeled, as described in the section titled *Model distribution system*"}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-6.png"))
```

@@

The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part. The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids. Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the duration of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multiyear trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature). The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited more modest changes in subsequent years that are difficult to attribute, but they may be a response to changes in water quality over the course of the study.

```{r figure-8, fig.cap="**Figure 8. (a) ** The global multiyear smooth and its estimated first derivative. A random sample of 200 smooths/derivatives is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions. The median of the global trend is shaded red when the 95% credible interval around the first derivative does not include zero. **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM. Ticks at the top and bottom edges of the plot in **(b)** represent anomalously high and left-censored iron concentrations, respectively."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-8.png"))
```

@@

Over the duration of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018].

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded the lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^. Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.

>>>>

All of the above present barriers to reliable statistical inference, but left-censoring and autocorrelation can be particularly challenging. Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data. And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].

@@

The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2). This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022]. GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

@@

To compare corrosion control treatments using data from the model distribution system, we fitted separate GAMs with continuous-time first-order autocorrelated errors to the particulate (>0.45 µm) and dissolved (<0.45 µm) lead time series. Particulate lead was determined as the difference between total and dissolved lead. Particulate lead concentrations less than the MDL of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored. The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration. Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were not modeled.

@@

The models included one-dimensional global smooths describing multiyear variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a). Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019]. The multiyear smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.

@@

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$.

@@

### Estimating rates of change

@@

The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers. Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset. Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation. The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.

@@

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term.

@@

```{r figure-6, fig.cap="**Figure 6.** Time series of dissolved and particulate lead representing three orthophosphate dosing protocols (columns) and two pipe configurations (rows; the full lead pipe configuration---Pb #1--2---is duplicated). The expected value of the posterior predictive distribution (*i.e.*, $\\mu_t$ in Equation 5) is summarized by the median (blue and red lines) and the 2.5--97.5^th^ percentiles (red/blue shaded regions). The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and ticks at the top of each panel represent values outside the plot limits. Other significant disturbances are labeled, as described in the section titled *Model distribution system*"}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-6.png"))
```

@@

The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or dissolved lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part. The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids [@trueman_characterizing_2019; @li_impact_2020]. Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multiyear trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature). The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the course of the study.

```{r figure-8, fig.cap="**Figure 8. (a) ** The global multiyear smooth and its estimated local slope. A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions. The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero. **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM. Ticks at the top and bottom edges of the plot in **(b)** represent anomalously high and left-censored iron concentrations, respectively."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-8.png"))
```

@@

Over the course of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018].

@@

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^. Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.
